The Ranked Nearest Neighbour Interchange space

### 2019-08-20 paper.tex
use alg:max_dist to prove radius=diameter

\begin{corollary}
    It is $\Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$.
\end{corollary}

@@

Besides the diameter, the \emph{radius} of a graph $G$, which is defined by $rad(G) =  \min\limits_u \max\limits_v d(u,v)$ is a further interesting property of a graph.
In Conjecture~\ref{conjecture:radius} we claim that the radius of $\rnni$ equals its diameter.
Algorithm~\ref{alg:max_dist_tree} suggests this, since it computes for an arbitrary input tree $T$ a caterpillar tree $R$ with distance $d(T,R) = \frac{(n-1)(n-2)}{2}$ for $n \leq k$. \todo{$k = ?$, check computationally (or prove this -> see conjecture below)}

\begin{conjecture}
    It is $rad(\rnni) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
    \label{conjecture:radius}
\end{conjecture}

@@

\begin{algorithm}[H]
\caption{MAX\_DISTANCE\_TREE($T$)}
\label{alg:max_dist_tree}
\begin{algorithmic}[1]
	\STATE $R$ only contains cherry taxa $a, b$ of $T$ and has internal node with rank $n-1$
    \STATE In list $L$ all taxa $\{1,\ldots,n\}\setminus\{a,b\}$ are sorted such that the ranks of their parents in $T$ do not decrease
	\FOR {taxon $a$ in $L$}
		\STATE Add an internal node as parent of $a$ on an edge in $R$ such that the new internal node has rank one less than the previously lowest internal node and $R$ is a caterpillar tree
	\ENDFOR
	\RETURN $R$
\end{algorithmic}
\end{algorithm}

\begin{conjecture}
    Let $T$ be an arbitrary tree and $R$ a caterpillar tree.
    Algorithm~\ref{alg:find_path} ($\findpath$) gives a path of length $\frac{(n-1)(n-2)}{2}$ between $T$ and $R$ if, and only if, $R$ can be generated from $T$ by applying Algorithm~\ref{alg:max_dist_tree}.
    \label{conjecture:max_dist_caterpillar}
\end{conjecture}

%TODO
%prove this conjecture! (<= would be sufficient for the conjecture above!)

>>>>

\begin{corollary}
    It is $\Delta(\rnniu) = \frac{(n-1)(n-2)}{2}$.
    \label{corollary:diameter}
\end{corollary}

@@

Besides the diameter, the \emph{radius} of a graph $G$, which is defined by $rad(G) =  \min\limits_u \max\limits_v d(u,v)$ is a further interesting property of a graph.
We will use Algorithm~\ref{alg:max_dist_tree} to show that in $\rnni$ the radius equals the diameter.

@@

\begin{algorithm}[H]
\caption{MAX\_DISTANCE\_TREE($T$)}
\label{alg:max_dist_tree}
\begin{algorithmic}[1]
	\STATE $R$ only contains cherry taxa $a, b$ of $T$ and has internal node with rank $n-1$
    \STATE In list $L$ all taxa $\{1,\ldots,n\}\setminus\{a,b\}$ are sorted such that the ranks of their parents in $T$ do not decrease
	\FOR {taxon $a$ in $L$}
		\STATE Add an internal node as parent of $a$ on an edge in $R$ such that the new internal node has rank one less than the previously lowest internal node
	\ENDFOR
	\RETURN $R$
\end{algorithmic}
\end{algorithm}

\begin{lemma}
    Let $T$ be a caterpillar tree.
    A tree $R$ that is computed by applying Algorithm~\ref{alg:max_dist_tree} on $R$ has distance $d(T,R) = \Delta(\rnni)$ to $T$.
    \label{lemma:max_dist_caterpillar}
\end{lemma}

\begin{proof}
    Recall that with Corollary~\ref{corollary:diameter} we know that $\Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
    Within this proof we assume, without loss of generality, that it is $T = (( \dots (1,2)3)\dots)n)$
    We prove the lemma by induction on the number of taxa $n$.

    Induction basis: $n = 3$\\
    Algorithm~\ref{alg:max_dist_tree} computes either $R = ((1,3),2)$ or $R'= ((2,3),1)$ when applied to $T = ((1,2),3)$.
    Since it is $d(T,R) = d(T,R') = 1 = \Delta(\rnni_3)$, the lemma is true for trees on $n=3$ taxa.

    Induction step: $n \to n+1$\\
    Induction hypothesis: For a caterpillar tree $T$ on $n$ taxa Algorithm~\ref{alg:max_dist_tree} computes a tree $R$ with distance $d(T,R) = \frac{(n-1)(n-2)}{2}$.\\
    Let us now  assume that $T$ is a tree on $n+1$ taxa.
    In the list $L$ produced in Algorithm~\ref{alg:max_dist_tree}, when applied to $T$, taxon $n+1$ is the last one.
    Hence $n+1$ is added to $R$ in the last step and therefore, it is in the lowest cherry of $R$ and has parent with rank $1$.
    Because the parent of $n+1$ has rank $n$ in $T$, we can conclude from Lemma~\ref{lemma:distance_delete_taxon} that for $T$ and $R$ restricted to taxa $1,\ldots,n$ it holds: $d(T|_n,R|_n) \leq d(T,R) - (n-1)$.
    It is easy to see that $R|_n$ can be received from $T|_n$ by applying Algorithm~\ref{alg:max_dist_tree}.
    So we can follow from the induction hypothesis and Lemma~\ref{lemma:distance_delete_taxon} that it is $d(T,R) \geq d(T|_n,R|_n) + (n-1) = \frac{(n-1)(n-2)}{2} = \frac{n(n-1)}{2}$, which concludes the proof of the lemma.
\end{proof}

In Algorithm~\ref{alg:max_dist_tree} it is not uniquely determined, where a new taxon is added to the already existing tree.
Instead, the attachment branch can be randomly chosen with the only restriction being that the rank of the new node needs to be below the ranks of already existing internal nodes.
This is the reason why one can receive every tree topology on $n$ taxa when applying Algorithm~\ref{alg:max_dist_tree} onto a caterpillar tree.
And because permuting the labels of the trees does not change the distance, there is for every tree $R$ on $n$ taxa a caterpillar tree $T$ with distance $d(T,R) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
This gives us Corollary~\ref{corollary:radius}.

\begin{corollary}
    The radius of the $\rnni$ graph is $rad(\rnni) = \Delta(\rnni) = \frac{(n-1)(n-2)}{2}$.
    \label{corollary:radius}
\end{corollary}
